%for a more compact document, add the option openany to avoid
%starting all chapters on odd numbered pages
\documentclass[12pt]{cmuthesis}

% This is a template for a CMU thesis.  It is 18 pages without any content :-)
% The source for this is pulled from a variety of sources and people.
% Here's a partial list of people who may or may have not contributed:
%
%        bnoble   = Brian Noble
%        caruana  = Rich Caruana
%        colohan  = Chris Colohan
%        jab      = Justin Boyan
%        josullvn = Joseph O'Sullivan
%        jrs      = Jonathan Shewchuk
%        kosak    = Corey Kosak
%        mjz      = Matt Zekauskas (mattz@cs)
%        pdinda   = Peter Dinda
%        pfr      = Patrick Riley
%        dkoes = David Koes (me)

% My main contribution is putting everything into a single class files and small
% template since I prefer this to some complicated sprawling directory tree with
% makefiles.

\definecolor{ValidRed}{rgb}{0.5,0,0}
\definecolor{TrueRed}{rgb}{1,0.6,0.2}
\definecolor{ValidBlue}{rgb}{0,0,0.5}
\definecolor{TrueBlue}{rgb}{0.2,0.6,1}

% some useful packages
\input{macros.tex}
%\draftstamp{\rm \today}{\rm DRAFT}

% \usepackage{titlesec,minitoc}  
% \titleclass{\part}{top}
% \titleformat{\part}
% {\centering\normalfont\Huge\bfseries} 
% {Foo} 
% {0pt} 
% {Bar}


\begin {document} 
\frontmatter

%initialize page style, so contents come out right (see bot) -mjz
\pagestyle{empty}

\title{ %% {\it \huge Thesis Proposal}\\
{\bf Substructural Logical Specifications}
%\\
%{\it (Fourth committee draft)}
}
\author{Robert J. Simmons}
\date{\today}
\Year{2013}
\trnumber{revision 1.0.1}

\committee{
Frank Pfenning, Chair \\
Robert Harper \\
Andr{\'e} Platzer \\
Iliano Cervesato, Carnegie Mellon Qatar \\
Dale Miller, INRIA-Saclay \& LIX/Ecole Polytechnique
}

\support{This research was sponsored by the National Science
  Foundation Graduate Research Fellowship; National Science Foundation
  under research grant number CNS-0716469; US Army Research Office
  under grant number DAAD-190210389; Funda\c{c}\~ao para a Ci\^encia e
  a Tecnologia (Portuguese Foundation for Science and Technology) via
  the Carnegie Mellon Portugal Program under grant NGN-44; X10
  Innovation Award from IBM; and Siebel Scholar Program.}  

\disclaimer{Any opinions, findings, conclusions or recommendations
  expressed in this publication are those of the author and do not
  necessarily reflect the views of any sponsoring institution or
  government.}

% copyright notice generated automatically from Year and author.
% permission added if \permission{} given.

\keywords{logical frameworks, linear logic, ordered logic, operational
  semantics}

\maketitle

% XXX MAKE DEDICATION
%\begin{dedication}
%To my grandparents
%\end{dedication}

\pagestyle{plain} % for toc, was empty

%% Obviously, it's probably a good idea to break the various sections of your thesis
%% into different files and input them into this file...

\begin{abstract}
  A logical framework and its implementation should serve as a
  flexible tool for specifying, simulating, and reasoning about formal
  systems. When the formal systems we are interested in exhibit state
  and concurrency, however, existing logical frameworks fall short of
  this goal. Logical frameworks based on a rewriting interpretation of
  substructural logics, ordered and linear logic in particular, can
  help. To this end, this dissertation introduces and demonstrates
  four methodologies for developing and using substructural logical
  frameworks for specifying and reasoning about stateful and
  concurrent systems.

  {\it Structural focalization} is a synthesis of ideas from
  Andreoli's focused sequent calculi and Watkins's hereditary
  substitution. We can use structural focalization to take a logic and
  define a restricted form of derivations, the {\it focused
    derivations}, that form the basis of a logical framework. We apply
  this methodology to define \sls, a logical framework for
  \underline{s}ubstructural \underline{l}ogical
  \underline{s}pecifications, as a fragment of ordered linear lax
  logic.

  {\it Logical correspondence} is a methodology for relating and
  inter-deriving different styles of programming language
  specification in \sls. The styles we connect range from very
  high-level specification styles like natural semantics, which do not
  fully specify the control structure of programs, to low-level
  specification styles like destination-passing, which provide
  detailed control over concurrency and control flow. We apply this
  methodology to systematically synthesize a low-level
  destination-passing semantics for a Mini-ML language extended with
  stateful and concurrent primitives. The specification is mostly
  high-level except for the relatively few rules that actually deal
  with concurrency.

  {\it Linear logical approximation} is a methodology for deriving
  program analyses by performing abstract analysis on the
  \sls~encoding of the language's operational semantics. We
  demonstrate this methodology by deriving a control flow analysis and
  an alias analysis from suitable programming language specifications.

  {\it Generative invariants} are a powerful generalization of
  both context-free grammars and LF's regular worlds that allow us
  to express invariants of \sls~specifications in \sls. We show that
  generative invariants can form the basis of
  progress-and-preservation-style reasoning about programming
  languages encoded in \sls.
\end{abstract}

% XXX MAKE ACKNOLWEDGEMENTS
\begin{acknowledgments}
  I cannot begin to fully acknowledge the debt of gratitude I owe to
  to my advisor, Frank Pfenning. Frank has provided me with
  extraordinary guidance from the moment I arrived in Wean Hall and
  started asking incredibly naive questions about these curious
  logical propositions $A \lolli B$ and ${!}A$. Nearly all of the
  ideas in this document were either developed or critically refined
  during our many conversations over the past seven years. I also want
  to thank my committee, Robert Harper, Andr{\'e} Platzer, Iliano
  Cervesato, and Dale Miller. Iliano's extremely timely and detailed
  feedback during the summer and fall of 2012 was especially important
  to making this final document as intelligible as it is.

  Many, many people have given me helpful feedback on this document
  and the various papers that went into it, especially Sean
  McLaughlin, Jason Reed, Rowan Davies, Chris Martens, Ian Zerny,
  Beniamino Accattoli, Jonathan Aldrich, Roger Wolff, Lindsey Kuper,
  and Ben Blum.  Beyond this group, the conversations I have had with
  Noam Zeilberger, Dan Licata, Bernardo Toninho, William Lovas,
  Carsten Sch{\"u}rmann, Henry DeYoung, Anders Schack-Nielsen, David
  Baelde, Taus Brock-Nannestad, Karl Crary, Carlo Angiuli, and both
  the Spring 2012 Linear Logic class and the CMU POP group as a whole
  were incredibly important as I developed the ideas that are
  presented here.  I'd also like to acknowledge the high-quality
  feedback I've received through anonymous review during my time in
  graduate school. Not everyone has a positive experience with peer
  review, and I appreciate the degree to which most of the
  reviews I've received have been both constructive and fair.

  And then there's the bigger picture.  So much of who I am is due to
  my family -- my extended family of grandparents, aunts, uncles, and
  cousins, but especially Mom, Dad, and Elizabeth. I would never have
  gotten to graduate school in the first place, nor made it through,
  without their unwavering love, support, and patience (so much
  patience).

  I have been fortunate to have the guidance of many exceptional
  academic and professional mentors along the way. I want to
  especially thank Jessica Hunt, Andrew Appel, David Walker, Sriram
  Rajamani, and Aditya Nori, who have each contributed unique and
  valuable pieces to my mathematical education.

  I also must thank the caring, quirky, and diverse Pittsburgh
  community that has helped me remember that there are things I care
  about above and beyond writing funny Greek symbols in notebooks and
  TeX files.  Rachel Gougian, of course, who stuck beside me for the
  whole ride -- I couldn't have done this without you, and even if I
  could have it wouldn't have been much fun. There was also Jen,
  Jamie, Pete, Laurie, Carsten, Nawshin, Tom, Jim, Mary, Thomas,
  Emily, Nick, Mike, and all my fellow Dec/5 troublemakers, espresso
  elves, talented Yinzerstars, SIGBOVIK organizers, and Words with
  Friends buddies.  

  It's been a privilege and a blessing. Thanks, everybody.

% COMMITTEE: Frank Robert Andr\`e Iliano Dale
% MENTORS (who aren't also committee):
%   - Jessica Hunt
%   - Andrew Appel, Dave Walker
%   - Sriram Rajamani, Aditya Nori
% DISCUSSERS (who aren't also readers):
%   - Noam Zeilberger
%   - Dan Licata
%   - The Linear Logic class
%   - Bernardo Toninho
%   - Henry De Young
%   - POP group, generically
% READERS: 
%   - Sean McLaughlin
%   - Jason Reed
%   - Rowan Davies
%   - Chris Martens 
%   - Ian Zerny
%   - Beniamino Accattoli
%   - Lindsey Kuper
%   - Ben Blum
%   - Jonathan Aldrich, Roger Wolff, and the three anonymous HOSC reviewers
%   - Three anonymous reviewers of the ICFP submission
%   - Five anonymous reviewers of the LICS submission
%   - Thank paper reviewers in general
% PEOPLE:
%   - Rachel, Jen, Jamie, Lauren, Pete, Laurie, Carsten, Nawshin, Mary
%   - Thomas, Emily, Nick, and the Yinzerstars
% FAMILY: Mom, Dad, Elizabeth
% WORDS WITH FRIENDS: Cyrus, Mom, Jonas, Chris, Ed
\end{acknowledgments}



\tableofcontents
\listoffigures % XXX Do I want this?
% \listoftables XXX Probably don't need this at all.

\mainmatter

%% Double space document for easy review:
%\renewcommand{\baselinestretch}{1.66}\normalsize

% The other requirements Catherine has:
%
%  - avoid large margins.  She wants the thesis to use fewer pages, 
%    especially if it requires colour printing.
%
%  - The thesis should be formatted for double-sided printing.  This
%    means that all chapters, acknowledgements, table of contents, etc.
%    should start on odd numbered (right facing) pages.
%
%  - You need to use the department standard tech report title page.  I
%    have tried to ensure that the title page here conforms to this
%    standard.
%
%  - Use a nice serif font, such as Times Roman.  Sans serif looks bad.
%
% Other than that, just make it look good...


% Introduction
\input{chapter-introduction}

\part{Focusing substructural logics}

% Linear logic
\input{chapter-foc}

% Substructural logic
\input{chapter-order}

% Substructural logical specifications
\input{chapter-framework}

\part{Substructural operational semantics}

% On logical correspondence
\input{chapter-correspondence}

% Ordered abstract machines
\input{chapter-absmachine}

% Destination-passing style
\input{chapter-destinations}

% Linear logical approximation
\input{chapter-approx}

\part{Reasoning about substructural logical specifications}

\input{chapter-gen}

\input{chapter-safety}

\input{chapter-conclusion}

\appendix

\input{appendix-procsum}

\input{appendix-hybrid}

% \chapter{Programming with canonical forms}

% \newcommand{\F}[1]{\ensuremath{F({#1})}}
% \newcommand{\G}[1]{\ensuremath{G(\textcolor{TrueBlue}{#1})}}
% \newcommand{\upX}[1]{\ensuremath{{\uparrow}\textcolor{ValidBlue}{#1}}}
% \newcommand{\downX}[1]{\ensuremath{{\downarrow}\textcolor{ValidRed}{#1}}}
% \newcommand{\upA}[1]{\ensuremath{{\uparrow}\textcolor{TrueBlue}{#1}}}
% \newcommand{\downA}[1]{\ensuremath{{\downarrow}\textcolor{TrueRed}{#1}}}

% \newcommand{\valid}[1]{\ensuremath{{\downarrow}\textcolor{ValidBlue}{{#1}\,\mathit{valid}}}}
% \newcommand{\true}[1]{\ensuremath{{\downarrow}\textcolor{TrueBlue}{{#1}\,\mathit{true}}}}

% \newcommand{\ajseq}[2]{\ensuremath{\mathstrut{#1} \vdash {#2}}}
% \newcommand{\ajinv}[3]{\ajseq{{#1}; {#2}}{\textcolor{ValidRed}{#3}}}
% \newcommand{\ajrfoc}[2]{\ajseq{{#1}}{[\textcolor{ValidBlue}{#2}]}}
% \newcommand{\ajlfoc}[3]{\ajseq{{#1} [{#2}]}{\textcolor{ValidRed}{#3}}}
% \newcommand{\ajAseq}[3]{\ensuremath{\mathstrut{#1} \vdash {#2}}}
% \newcommand{\ajAinv}[4]{\ajseq{{#1}; {#2}}{\textcolor{ValidRed}{#3}}}
% \newcommand{\ajArfoc}[3]{\ajseq{{#1}}{[\textcolor{ValidBlue}{#2}]}}
% \newcommand{\ajAlfoc}[4]{\ajseq{{#1} [{#2}]}{\textcolor{ValidRed}{#3}}}
% \newcommand{\ajXseq}[2]{\ensuremath{\mathstrut{#1} \vdash {#2}}}
% \newcommand{\ajXinv}[3]{\ajseq{{#1}; {#2}}{\textcolor{ValidRed}{#3}}}
% \newcommand{\ajXrfoc}[2]{\ajseq{{#1}}{[\textcolor{ValidBlue}{#2}]}}
% \newcommand{\ajXlfoc}[3]{\ajseq{{#1} [{#2}]}{\textcolor{ValidRed}{#3}}}

% \begin{figure}
% \fbox{\ajXrfoc{\Gamma}{X^+}}
% \[
% \infer[x^+_R]
% {\ajrfoc{\valid{x^+}}{x^+}}
% {}
% \qquad
% \infer[{\downarrow}_{XR}]
% {\ajXrfoc{\Gamma}{\downX{X^-}}}
% {\ajXinv{\Gamma}{\cdot}{X^-}}
% \qquad
% \infer[G_R]
% {\ajXrfoc{\Gamma}{\G{A}}}
% {\ajArfoc{\Gamma}{\cdot}{A}}
% \]

% \fbox{\ajArfoc{\Gamma}{\Delta}{A^+}}
% \[
% \infer[a^+_R]
% {\ajArfoc{\cdot}{a^+}{a^+}}
% {}
% \qquad
% \infer[{\downarrow}_{AR}]
% {\ajArfoc{\Gamma}{\Delta}{\downA{A^-}}}
% {\ajAinv{\Gamma}{\Delta}{\cdot}{A^-}}
% \]

% \caption{Focused adjoint logic}
% \end{figure}


%\include{appendix}

\backmatter

%\renewcommand{\baselinestretch}{1.0}\normalsize

% By default \bibsection is \chapter*, but we really want this to show
% up in the table of contents and pdf bookmarks.
\renewcommand{\bibsection}{\chapter{\bibname}}
%\newcommand{\bibpreamble}{This text goes between the ``Bibliography''
%  header and the actual list of references}
\bibliographystyle{alphaurl}
\bibliography{ref} %your bib file

\end{document}
